Thesis Title

### 2022-03-26 chapters/acconyms.tex
update

\begin{acronym}[MPC] % Give the longest label here so that the list is nicely aligned
\acro{GBD}{Global Burden of Disease}
\acro{WHO}{World Health Organization}
\acro{SARS-CoV-2}{Acute  Respiratory  Syndrome  Coronavirus  2}
\acro{MERS}{Middle East Respiratory Syndrome}
\acro{GPP}{Global Priority Pathogens}
\acro{AMR}{Antimicrobial Resistance}

@@

T2 Magnetic Resonance (T2MR)
polymerase chain reaction (PCR)
pulsed field gel electrophoresis (PFGE)
Multilocus sequence typing (MLST)
basepairs (bp)
Real-time quantitative PCR (qPCR)
Lateral flow immunoassays (LFIA)
ribossomal RNA (rRNA)
PubMed Central\textsuperscript{\small\textregistered} (PMC)
Public health surveillance systems (PHSS)
dideoxynucleotide triphosphate (ddNTP) 
deoxynucleotide triphosphate (dNTP)
single molecule real-time sequencing (SMRT)
Whole genome sequencing (WGS)
ribosomal RNA (rRNA)
Operational Taxonomic Units (OTUs)
Shiga toxin-producing \textit{Escherichia coli} (STEC)
National Institute of Standards and Technology (NIST)
reverse transcription-PCR (RT-PCR)
core-genome/whole genome MLST (cg/wg MLST)
overlap-layout consensus (OLC)
Cerebrospinal fluid (CSF)

>>>>

\begin{acronym}[MPC] % Give the longest label here so that the list is nicely aligned
\acro{GBD}{Global Burden of Disease}
\acro{WHO}{World Health Organization}
\acro{SARS-CoV-2}{Acute  Respiratory  Syndrome  Coronavirus  2}
\acro{MERS}{Middle East Respiratory Syndrome}
\acro{GPP}{Global Priority Pathogens}
\acro{AMR}{Antimicrobial Resistance}
\acro{PCR}{Polymerase Chain Reaction}
\acro{PFGE}{Pulse Field Gel Electrophoresis}
\acro{MLST}{Multilocus Sequence Typing}

@@

basepairs (bp)
Real-time quantitative PCR (qPCR)
Lateral flow immunoassays (LFIA)
ribossomal RNA (rRNA)
PubMed Central\textsuperscript{\small\textregistered} (PMC)
Public health surveillance systems (PHSS)
dideoxynucleotide triphosphate (ddNTP) 
deoxynucleotide triphosphate (dNTP)
single molecule real-time sequencing (SMRT)
Whole genome sequencing (WGS)
ribosomal RNA (rRNA)
Operational Taxonomic Units (OTUs)
Shiga toxin-producing \textit{Escherichia coli} (STEC)
National Institute of Standards and Technology (NIST)
reverse transcription-PCR (RT-PCR)
core-genome/whole genome MLST (cg/wg MLST)
overlap-layout consensus (OLC)
Cerebrospinal fluid (CSF)
core-genome/whole genome MLST (cg/wg MLST)

### 2022-03-26 chapters/introduction.tex
update

The past few decades have seen a major revolution in the operation of microbial laboratories, driven by the development of molecular technologies and ways to make these accessible, namely amplification-based polymerase chain reaction (PCR), matrix-assisted laser desorption/ionisation - time of flight (MALDI-TOF) and DNA-microarray-based hybridisation technology. 
These are used in conjunction with traditional techniques such as microscopy, culture, and serology.
Application of these methods differs by suspected infection type: bacterial, viral, fungal or parasitic. 
For the purpose of this dissertation work, we will focus on bacterial (see Section \ref{sssec:_intro_bacterial} and viral infections (see Section \ref{sssec:_intro_viral}.

@@

The current gold standard for bacterial pathogen identification in diagnostic microbiology laboratories involves the isolation of the pathogen through culture followed by biochemical testing, a multi-step process that can take days to weeks before obtaining results, depending on the fastidiousness of the organism and if it can be cultured \citep{muhamad_rizal_advantages_2020, giuliano_guide_2019, muhamad_rizal_advantages_2020}. 
Although culture allows the identification of a wide variety of organisms, some pathogens can escape routine investigation due to strict metabolic necessities for growth or the requirement for specific biochemical tests needed for their identification. 
Furthermore, results will be obscured if a mixed culture is obtained, particularly if the cultures are obtained from sites with a microbiota, such as the gut and the skin, increasing the risk of contamination by normal flora and leading to false results \citep{giuliano_guide_2019}. 
After successful growth in culture, Gram staining and MALDI-TOF mass spectrometry are often used for identification with good accuracy as long as the pathogen is presented in the the coexisting database \citep{patel_maldi-tof_2015}. 
An alternate rapid identification method is PCR where nucleic acid fragments are detected through specific primers, being highly sensitive and specific, to the point where PCR may detect bacteria that are not viable after a patient has been treated for an infection and it is limited to the primer used \citep{scerbo_beyond_2016}. 
Syndromic panels, an extension of PCR by using multiple primers (multiplex PCR) to simultaneously amplify nucleic acids from multiple targets in a single reaction, tried to address this issue by allowing for the identification of multiple bacteria and other important information such as the detection of antibiotic resistance or virulence genes \citep{giuliano_guide_2019}.

Following identification, antibiotic-susceptibility testing is essential to guide clinicians in selecting an appropriate treatment. 
Conventional methods of bacterial resistance detection, such as disc diffusion, antimicrobial gradient strip, and broth microdilution, are widely used, but results cannot be obtained before 48 hours after receiving a sample, which can lead to prolonged use or overuse of broad-spectrum antibiotics \citep{benkova_antimicrobial_2020}. 
Similarly to bacterial identification, MALDI-TOF and PCR have been increasingly adopted as solutions with shorter turnaround times, although no phenotypic information is recovered, nor information on the minimum inhibitory concentration (MIC) for a given antibiotic.

Choosing an appropriate bacterial typing technique for epidemiological studies depends on the available resources and the minimum intended resolution, ranging from DNA fingerprinting to multilocus sequence typing, Pulsed-field gel electrophoresis (PFGE), and sequence-based typing (see \secref{sec:_intro_genomics_approach}) \citep{allerberger_molecular_2012,foxman_choosing_2005}. 
DNA macrorestriction analysis by PFGE, which revolutionised precise separation of DNA fragments, became the most widely implemented DNA fingerprinting technique \citep{allerberger_molecular_2012}, becoming the golden standard for bacterial typing \citep{neoh_pulsed-field_2019}.

In the early 2000s, Multilocus sequence typing (MLST) was proposed as a portable, universal, and definitive method for characterising bacteria \citep{maiden_multilocus_2006}. 
Instead of enzyme restriction of bacteria DNA, separation of restricted DNA bands using a PFGE chamber, followed by clonal assignment of bacteria based on banding patterns, MLST relies on the amplification through PCR sequences of internal fragments of housekeeping genes (usually 5 to 7), approximately 450-500 basepairs (bp) in size, followed by its sequence, usually my Sanger methods (see \secref{ssec:_intro_1st_gen_seq}). 
For each house-keeping gene, the different sequences present within a bacterial species are assigned as distinct alleles and, for each isolate, the alleles at each of the (usually) seven loci define the allelic profile or sequence type \citep{larsen_multilocus_2012}. 
As with PFGE, different schemes, defining what housekeeping gene fragments are used, are available depending on the species. 
Unlike PFGE, the provision of freely accessible, curated databases of MLST nucleotide sequence data enables the direct comparison of bacterial isolates, providing the basis of a common language for bacterial typing \citep{maiden_multilocus_2006}. 
So far, MLST schemes for more than 100 bacterial organisms have been published and made freely available\footnote{\url{https://pubmlst.org/organisms}}, \cite{jolley_open-access_2018})

@@

Direct detection methods negate the need for virus propagation, detecting the virus directly from the suspect source through nucleic acid and immunological methods. 
PCR and reverse transcription-PCR (RT-PCR) are widely applied methods for the detection of both DNA and RNA viruses, respectively, driven by increased awareness of the clinical value of and demand for prompt information about viral loads, viral sequence data and potential antiviral resistance information \citep{cassedy_virus_2021}. 
Syndromic testing (see \secref{sssec:_intro_bacterial}) is now fully integrated into standard testing practises of many clinical laboratories \citep{dien_bard_panels_2020}. 
The limitations of these assays include the absence of detection of off-target pathogens, a lack of full susceptibility information, cost, and false positive results. 
Real-time quantitative PCR (qPCR) remains the front line tool in aetiological diagnosis, measuring the production of the target amplicon throughout the reaction and providing quantitative results with high specificity and sensibility, albeit with a significant cost due to sophisticated apparatus despite high-throughput systems being widely established \citep{cassedy_virus_2021}.

@@

The first generation sequencing technology remains very useful for applications where high-throughput is not required due to its cost-effectiveness, relatively low sample load and accuracy of sequencing, even in repetitive genomic regions, although input DNA must consist of a relatively pure population of sequences \citep{slatko_overview_2018}. 
One of the most common uses is thus individual sequencing reactions using a specific DNA primer on a specific template, such as MLST of bacterial genomes.

@@

Despite the increasing adoption of DNA sequencing methods in clinical microbiology, the sequencing of genetic material from a pure culture requires \textit{a priori} knowledge of what to expect from a particular clinical sample or patient \citep{schuele_future_2021}. 
In most cases, this knowledge is enough to request the most appropriate test, such as multiplexed panels or specific culture media, but this is not always the case. 
In recent years, there has been a growing interest in using metagenomics to deliver culture-independent approaches to microbial ecology, surveillance and diagnosis (see Figure \ref{fig:figure4})\citep{loman_twenty_2015, loman_high-throughput_2012}.
Metagenomic DNA sequence allows detailed characterisation of pathogens in all kinds of samples originating from humans, animals, food and the environment, ligating the diagnostics to surveillance in a true "one health" fashion \citep{rossen__2018}. 
Unlike PCR or microarrays, it usually does not require primer or probe design, it can be easily multiplexed, and the specificity and selectivity of the sequencing can be adjusted computationally after acquiring the data \citep{dunne_next-generation_2012}.  
While most molecular assays target only a limited number of pathogens, metagenomic approaches characterise all DNA or RNA present in a sample, enabling analysis of the entire microbiome as well as the human host genome or transcriptome in patient samples \citep{chiu_clinical_2019}. 
Whether or not it can entirely replace routine microbiology depends on several conditions and future developments, both technological and computational (see \secref{sec:_intro_bioinformatics}).

@@

Because this approach is PCR-based, it suffers from the same issues described previously for conventional PCR, requiring primer design. 
Additionally, it must necessarily account for intragenomic variation between 16S gene copies. 
Microbial profiles generated using different primer pairs need independent validation of performance, and the comparison of data sets across V-regions using different databases might be misleading due to differences in nomenclature and varying precisions in classification, and specific but important taxa are not picked up by certain primer pairs (e.g., \textit{Bacteroidetes} is missed using primers 515F-944R) or due to the database used \citep{abellan-schneyder_primer_2021}. 
Furthermore, targeting of 16S variable regions with short-read sequencing platforms cannot achieve the taxonomic resolution afforded by sequencing the entire (~1500 bp) gene \citep{johnson_evaluation_2019}. 
The emergence of third generating sequencing technologies (see \secref{ssec:_intro_3rd_gen_seq}) allows for this limitation to be overcome but currently, only a fraction of the databases includes complete 16S rRNA sequences.

@@

The last step on the assembly methodology is the evaluation of the completeness and contamination of the bins. When using a taxonomic binner, the effects of contamination are mitigated as the sequence clustering is performed based on matches with reference database. The contaminants, if present in the database, will be separated into different bins or just added to the bin of unclassified sequences. When using a taxonomic independent binning software, the composition and abundance might not be enough to discriminate between all the organisms, with the possible result of having bins with contaminating sequences of other organisms present in the sample. CheckM\footnote{\url{https://ecogenomics.github.io/CheckM/}} assesses the quality of the recovered genomes, estimating completeness and contamination by evaluating ubiquitous single-copy genes \citep{parks_checkm_2015}.

>>>>

The past few decades have seen a major revolution in the operation of microbial laboratories, driven by the development of molecular technologies and ways to make these accessible, namely amplification-based \ac{PCR}, matrix-assisted laser desorption/ionisation - time of flight (MALDI-TOF) and DNA-microarray-based hybridisation technology. 
These are used in conjunction with traditional techniques such as microscopy, culture, and serology.
Application of these methods differs by suspected infection type: bacterial, viral, fungal or parasitic. 
For the purpose of this dissertation work, we will focus on bacterial (see Section \ref{sssec:_intro_bacterial} and viral infections (see Section \ref{sssec:_intro_viral}.

@@

The current gold standard for bacterial pathogen identification in diagnostic microbiology laboratories involves the isolation of the pathogen through culture followed by biochemical testing, a multi-step process that can take days to weeks before obtaining results, depending on the fastidiousness of the organism and if it can be cultured \citep{muhamad_rizal_advantages_2020, giuliano_guide_2019, muhamad_rizal_advantages_2020}. 
Although culture allows the identification of a wide variety of organisms, some pathogens can escape routine investigation due to strict metabolic necessities for growth or the requirement for specific biochemical tests needed for their identification. 
Furthermore, results will be obscured if a mixed culture is obtained, particularly if the cultures are obtained from sites with a microbiota, such as the gut and the skin, increasing the risk of contamination by normal flora and leading to false results \citep{giuliano_guide_2019}. 
After successful growth in culture, Gram staining and MALDI-TOF mass spectrometry are often used for identification with good accuracy as long as the pathogen is presented in the the coexisting database \citep{patel_maldi-tof_2015}. 
An alternate rapid identification method is \ac{PCR} where nucleic acid fragments are detected through specific primers, being highly sensitive and specific, to the point where \ac{PCR} may detect bacteria that are not viable after a patient has been treated for an infection and it is limited to the primer used \citep{scerbo_beyond_2016}. 
Syndromic panels, an extension of \ac{PCR} by using multiple primers (multiplex \ac{PCR}) to simultaneously amplify nucleic acids from multiple targets in a single reaction, tried to address this issue by allowing for the identification of multiple bacteria and other important information such as the detection of antibiotic resistance or virulence genes \citep{giuliano_guide_2019}.

Following identification, antibiotic-susceptibility testing is essential to guide clinicians in selecting an appropriate treatment. 
Conventional methods of bacterial resistance detection, such as disc diffusion, antimicrobial gradient strip, and broth microdilution, are widely used, but results cannot be obtained before 48 hours after receiving a sample, which can lead to prolonged use or overuse of broad-spectrum antibiotics \citep{benkova_antimicrobial_2020}. 
Similarly to bacterial identification, MALDI-TOF and \ac{PCR} have been increasingly adopted as solutions with shorter turnaround times, although no phenotypic information is recovered, nor information on the minimum inhibitory concentration (MIC) for a given antibiotic.

Choosing an appropriate bacterial typing technique for epidemiological studies depends on the available resources and the minimum intended resolution, ranging from DNA fingerprinting to multilocus sequence typing, \ac{PFGE}, and sequence-based typing (see \secref{sec:_intro_genomics_approach}) \citep{allerberger_molecular_2012,foxman_choosing_2005}. 
DNA macrorestriction analysis by \ac{PFGE}, which revolutionised precise separation of DNA fragments, became the most widely implemented DNA fingerprinting technique \citep{allerberger_molecular_2012}, becoming the golden standard for bacterial typing \citep{neoh_pulsed-field_2019}.

In the early 2000s, \ac{MLST} was proposed as a portable, universal, and definitive method for characterising bacteria \citep{maiden_multilocus_2006}. 
Instead of enzyme restriction of bacteria DNA, separation of restricted DNA bands using a \ac{PFGE} chamber, followed by clonal assignment of bacteria based on banding patterns, \ac{MLST} relies on the amplification through \ac{PCR} sequences of internal fragments of housekeeping genes (usually 5 to 7), approximately 450-500 basepairs (bp) in size, followed by its sequence, usually my Sanger methods (see \secref{ssec:_intro_1st_gen_seq}). 
For each house-keeping gene, the different sequences present within a bacterial species are assigned as distinct alleles and, for each isolate, the alleles at each of the (usually) seven loci define the allelic profile or sequence type \citep{larsen_multilocus_2012}. 
As with \ac{PFGE}, different schemes, defining what housekeeping gene fragments are used, are available depending on the species. 
Unlike \ac{PFGE}, the provision of freely accessible, curated databases of \ac{MLST} nucleotide sequence data enables the direct comparison of bacterial isolates, providing the basis of a common language for bacterial typing \citep{maiden_multilocus_2006}. 
So far, \ac{MLST} schemes for more than 100 bacterial organisms have been published and made freely available\footnote{\url{https://pubmlst.org/organisms}}, \cite{jolley_open-access_2018})

@@

Direct detection methods negate the need for virus propagation, detecting the virus directly from the suspect source through nucleic acid and immunological methods. 
\ac{PCR} and reverse transcription-PCR (RT-PCR) are widely applied methods for the detection of both DNA and RNA viruses, respectively, driven by increased awareness of the clinical value of and demand for prompt information about viral loads, viral sequence data and potential antiviral resistance information \citep{cassedy_virus_2021}. 
Syndromic testing (see \secref{sssec:_intro_bacterial}) is now fully integrated into standard testing practises of many clinical laboratories \citep{dien_bard_panels_2020}. 
The limitations of these assays include the absence of detection of off-target pathogens, a lack of full susceptibility information, cost, and false positive results. 
Real-time quantitative PCR (qPCR) remains the front line tool in aetiological diagnosis, measuring the production of the target amplicon throughout the reaction and providing quantitative results with high specificity and sensibility, albeit with a significant cost due to sophisticated apparatus despite high-throughput systems being widely established \citep{cassedy_virus_2021}.

@@

The first generation sequencing technology remains very useful for applications where high-throughput is not required due to its cost-effectiveness, relatively low sample load and accuracy of sequencing, even in repetitive genomic regions, although input DNA must consist of a relatively pure population of sequences \citep{slatko_overview_2018}. 
One of the most common uses is thus individual sequencing reactions using a specific DNA primer on a specific template, such as \ac{MLST} of bacterial genomes.

@@

Despite the increasing adoption of DNA sequencing methods in clinical microbiology, the sequencing of genetic material from a pure culture requires \textit{a priori} knowledge of what to expect from a particular clinical sample or patient \citep{schuele_future_2021}. 
In most cases, this knowledge is enough to request the most appropriate test, such as multiplexed panels or specific culture media, but this is not always the case. 
In recent years, there has been a growing interest in using metagenomics to deliver culture-independent approaches to microbial ecology, surveillance and diagnosis (see Figure \ref{fig:figure4})\citep{loman_twenty_2015, loman_high-throughput_2012}.
Metagenomic DNA sequence allows detailed characterisation of pathogens in all kinds of samples originating from humans, animals, food and the environment, ligating the diagnostics to surveillance in a true "one health" fashion \citep{rossen__2018}. 
Unlike \ac{PCR} or microarrays, it usually does not require primer or probe design, it can be easily multiplexed, and the specificity and selectivity of the sequencing can be adjusted computationally after acquiring the data \citep{dunne_next-generation_2012}.  
While most molecular assays target only a limited number of pathogens, metagenomic approaches characterise all DNA or RNA present in a sample, enabling analysis of the entire microbiome as well as the human host genome or transcriptome in patient samples \citep{chiu_clinical_2019}. 
Whether or not it can entirely replace routine microbiology depends on several conditions and future developments, both technological and computational (see \secref{sec:_intro_bioinformatics}).

@@

Additionally, it must necessarily account for intragenomic variation between 16S gene copies. 
Microbial profiles generated using different primer pairs need independent validation of performance, and the comparison of data sets across V-regions using different databases might be misleading due to differences in nomenclature and varying precisions in classification, and specific but important taxa are not picked up by certain primer pairs (e.g., \textit{Bacteroidetes} is missed using primers 515F-944R) or due to the database used \citep{abellan-schneyder_primer_2021}. 
Furthermore, targeting of 16S variable regions with short-read sequencing platforms cannot achieve the taxonomic resolution afforded by sequencing the entire (~1500 bp) gene \citep{johnson_evaluation_2019}. 
The emergence of third generating sequencing technologies (see \secref{ssec:_intro_3rd_gen_seq}) allows for this limitation to be overcome but currently, only a fraction of the databases includes complete 16S rRNA sequences.

@@
